---
title: "R Notebook"
output: rmarkdown::github_document
---


Load libraries 

```{r libraries, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
library(glue)
library(R.utils)
library(janitor)
library(caret)
library(reshape2)
library(gridExtra)
library(grid)
library(lattice)
library(skimr)
library(ggthemr)
```

Load data, set ggplot theme

```{r}
#gunzip("C:/Users/sgran/Downloads/covtype.data.gz")

raw_data <- read_csv("C:/Users/sgran/Downloads/covtype.data",
                     col_names = FALSE)

ggthemr('fresh')
```

Format column names

```{r}
col_names <- c(
  "elevation",
  "aspect",
  "slope",
  "horizontal_distance_to_hydrology",
  "vertical_distance_to_hydrology",
  "horizontal_distance_to_roadways",
  "hillshade_9am",
  "hillshade_noon",
  "hillshade_3pm",
  "horizontal_distance_to_fire_points",
  glue::glue("wilderness_area{x}", x = 1:4),
  glue::glue("soil_type{x}", x = 1:40),
  "cover_type"
)

names(raw_data) <- col_names

raw_data$cover_type <- factor(
  x = raw_data$cover_type,
  levels = 1:7,
  labels = c(
    "spruce_fir",
    "lodgepole_pine",
    "ponderosa_pine",
    "cottonwood_Willow",
    "aspen",
    "douglasfir",
    "krummholz"
  )
)
```

```{r}
raw_data %>% tabyl(cover_type)

```

Split data

```{r}
set.seed(10)

train_index <- caret::createDataPartition(
  y = raw_data$cover_type,
  p = 0.7,
  times = 1,
  list = FALSE
)
```

```{r}
raw_train <- raw_data[train_index,]
raw_test  <- raw_data[-train_index,]
```

Look at training data, high level skim

```{r}
skim(raw_train)
```

Check soil variables
What soil is present for different cover types?

```{r}
train_soil <- raw_train %>%
    dplyr::select(cover_type, contains("soil")) %>%
    gather(-cover_type, key = soil, value = val) %>%
    mutate(soil = parse_number(soil), n=1) 

train_soil %>% 
    group_by(cover_type, soil) %>%
    summarise(total = sum(val),
              prop  = total / sum(n)) %>% 
    ggplot(aes(soil, prop, fill = cover_type)) + 
    geom_col() +
    facet_grid(~cover_type) +
    coord_flip()
```

Proportion of soils present for each cover type

```{r}
train_soil %>% 
    filter(val > 0) %>% 
    ggplot(aes(cover_type, val, fill = soil)) +
    geom_col(position="fill") +
    coord_flip()
```

Correlations of soil types

```{r}
raw_train %>% 
    dplyr::select(contains("soil")) %>% 
    cor() %>% 
    melt() %>% 
    mutate(v1 = parse_number(Var1),
           v2 = parse_number(Var2)) %>% 
    ggplot(aes(v1, v2, fill = value)) +
    geom_tile()

```

Check wilderness variables
What wilderness is there with each cover type?

```{r}
train_wild <- raw_train %>%
    dplyr::select(cover_type, contains("wilderness")) %>%
    gather(-cover_type, key = wild, value = val) %>%
    mutate(wild = parse_number(wild), n=1) 

train_wild %>% 
    group_by(cover_type, wild) %>%
    summarise(total = sum(val),
              prop  = total / sum(n)) %>% 
    ggplot(aes(wild, prop, fill = cover_type)) + 
    geom_col() +
    facet_grid(~cover_type) +
    coord_flip()
```

Proportion of wilderness area for each cover type

```{r}
train_wild %>% 
    filter(val > 0) %>% 
    ggplot(aes(cover_type, val, fill = wild)) +
    geom_col(position="fill") +
    coord_flip()
```

Wilderness area correlations

```{r}
raw_train %>% 
    dplyr::select(contains("wilderness")) %>% 
    cor() %>% 
    melt() %>% 
    mutate(v1 = parse_number(Var1),
           v2 = parse_number(Var2)) %>% 
    ggplot(aes(v1, v2, fill = value)) +
    geom_tile()
```

Proportion of soil/widerness combos for
each cover type

```{r}
raw_train %>% 
    dplyr::select(cover_type, contains("soil"), contains("wilderness")) %>% 
    mutate(n=1) %>% 
    gather(contains("soil"), key = soil, value = soil_val) %>%
    gather(contains("wilderness"), key = wild, value = wild_val) %>%
    filter(soil_val > 0, wild_val > 0) %>% 
    group_by(cover_type, soil, wild) %>% 
    summarise(n = sum(n)) %>% 
    ungroup() %>% 
    group_by(cover_type) %>% 
    mutate(prop = n / sum(n)) %>%
    mutate(soil = parse_number(soil), 
           wild = parse_number(wild)) %>% 
    ungroup() %>% 
    ggplot(aes(wild, soil, color=cover_type, size = prop)) + 
    geom_point() +
    facet_grid(~cover_type) 
```

Shade measure by time of day
for each cover type

```{r}
raw_train %>% 
    dplyr::select(cover_type, contains("hillshade")) %>% 
    gather(-cover_type, key = shade, value = val) %>% 
    mutate(shade = factor(shade, levels = c("hillshade_9am", "hillshade_noon", "hillshade_3pm"))) %>% 
    ggplot(aes(cover_type, val)) +
    geom_jitter(aes(color = shade), alpha = 0.05, stroke = 0) +
    geom_boxplot(aes(fill = shade)) +
    coord_flip() +
    facet_grid(~shade) +
    theme(legend.position="none")
```

Distance measures for each cover type

```{r}
raw_train %>% 
    dplyr::select(cover_type, contains("distance")) %>% 
    gather(-cover_type, key = dist, value = val) %>% 
    ggplot(aes(cover_type, val)) +
    geom_jitter(aes(color = dist), alpha = 0.05, stroke = 0) +
    geom_boxplot(aes(fill = dist)) +
    coord_flip() +
    facet_grid(~dist, scales = "free") +
    theme(legend.position="none")
```

```{r}
raw_train %>% 
    dplyr::select(cover_type, contains("hydrology")) %>% 
    ggplot(aes(horizontal_distance_to_hydrology, vertical_distance_to_hydrology)) +
    geom_point(aes(color = cover_type), alpha = 0.01, stroke = 0) +
    geom_smooth() +
    facet_grid(~cover_type) +
    theme(legend.position="none")
```

Remaining variables for each cover type

```{r}
raw_train %>% 
    dplyr::select(cover_type, aspect, elevation, slope) %>% 
    gather(-cover_type, key = key, value = val) %>% 
    ggplot(aes(cover_type, val, fill = cover_type)) +
    geom_boxplot() +
    coord_flip() +
    facet_grid(~key, scales = "free") +
    theme(legend.position="none")
```
